Likelihoods for a general class of ARGs under the SMC

### 2025-04-29 review-responses.tex

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are writing to submit a revised version of our manuscript,
GENETICS-2025-307907 
titled
``Likelihoods for a general class of ARGs under the SMC''.
We appreciate the thoughtful comments by the reviewers.
We've done our best to address all the comments,
as detailed in our Response to Reviewers.

@@

\begin{point}{\revref} % 300:
 cite what simplification does (Wang et al., YYYY; Kelleher et al. - your PLOS Genetics paper). Do you need to cite this sooner in the manuscript?
\end{point}

\reply{
Done.
}

\begin{point}{\revref} % 300:
 for this application, despite retaininig fewer nodes and edges.
\end{point}

\reply{
We would prefer to keep this sentence as it is, as ``retaining" here could 
be confusing --- the calculation doesn't keep or retain nodes and edges, that is a 
property of the input ARG.
}

### 2025-04-29 paper.tex
Batch of responses

% % % % % % % % % % % % % % % % % %
\subsection*{Robustness to partial and incorrect information}

Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any} \revpoint{2}{29}
collection of nodes (ancestral haplotypes that lived at some time
$t$ in the past) and edges (a record that child node $c$
inherited the genomic interval $[x, y)$ from parent node $p$)
under the SMC. Other than some basic consistency constraints (e.g., parents
must be older than children) there are no requirements on the topology
described, and in particular, no requirement that the events from the
model (the SMC) should be explicitly identified in the graph. Our
formulation reasons about the events that \emph{could have resulted in}
the observed nodes and edges, and therefore frees us from the requirement
of having to explicitly write those down. In this section we
illustrate the robustness
of this approach to incomplete and potentially incorrect
information about recombination and
coalescent events via a proof-of-concept example.
% in which we
% remove various levels of detail from a simulated ARG.

@@

In Fig.~\ref{fig:lik-surface} we begin with a realisation of the SMC
containing complete information about every simulated event
(Full ARG; 9,378 nodes; 13,221 edges; 2,933 trees). \revpoint{2}{30}
We then plot the likelihood computed by Eq.~\eqref{eq:full-lik}
as we vary the recombination rate parameter $r$, as a ratio
of the likelihood at the true value of $r$.
Reassuringly, we can
see that the minimum of the likelihood-ratio curve when using the
full simulated ARG is close to the true value of the parameter.
(The likelihood ratio curve obtained when computing the likelihood
under the full coalescent with recombination model using the KYF
formula is identical; not shown.)
We next compute the likelihood ratio for the ARG obtained when we
simplify out the recombination nodes from the full ARG
(``RE nodes removed''; 3,332 nodes; 6,901 edges; 2,933 trees; note that msprime
uses distinct nodes for the left and right parent of each recombination).
The curve we obtain is indistinguishable from that of the full
ARG, demonstrating that the locally unary portions of coalescent nodes
contain all the information that we need about recombination events
for this application.
When we fully simplify the simulated ARG
to remove all locally unary portions of nodes from trees
(``Fully simplified''; 3,332 nodes; 10,674 edges, 2,933 trees)  \revpoint{2}{31}
we can see there is a significant loss of information, \revpoint{2}{32}
and the minimum of the
likelihood ratio surface is biased away from its true value,
as expected, because removing the unary portions of nodes has removed some of the area
of the ARG in which recombination did not occur.
(See~\cite{wong_general_2023} for more discussion on varying
degrees of ARG simplification.)

>>>>

% % % % % % % % % % % % % % % % % %
\subsection*{Robustness to partial and incorrect information}
The ARG likelihood formulation in 
Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for any \revpoint{2}{29}
collection of nodes (haplotypes that existed at some time $t$) and 
edges (a record that child node $c$
inherited the genomic interval $[x, y)$ from parent node $p$)
under the SMC. Other than some basic consistency constraints (e.g., parents
must be older than children) there are no requirements on the topology
described, and in particular, no requirement that the events from the
model (the SMC) should be explicitly identified in the graph. Our
formulation reasons about the events that \emph{could have resulted in}
the observed nodes and edges, and therefore frees us from the requirement
of having to explicitly write those down. In this section we
illustrate the robustness
of this approach to incomplete and potentially incorrect
information about recombination and
coalescent events via a proof-of-concept example.
% in which we
% remove various levels of detail from a simulated ARG.

@@

In Fig.~\ref{fig:lik-surface} we begin with 
with a sample from the SMC model 
containing complete information about every simulated event
(Full ARG; 9,378 nodes; 13,221 edges; 2,933 trees). \revpoint{2}{30}
We then plot the likelihood computed by Eq.~\eqref{eq:full-lik}
as we vary the recombination rate parameter $r$, as a ratio
of the likelihood at the true value of $r$.
Reassuringly, we can
see that the minimum of the likelihood-ratio curve when using the
full simulated ARG is close to the true value of the parameter.
(The likelihood ratio curve obtained when computing the likelihood
under the full coalescent with recombination model using the KYF
formula is identical; not shown.)
We next compute the likelihood ratio for the ARG obtained when we
simplify~\citep{kelleher_efficient_2018,wong_general_2023}
 out the recombination nodes from the full ARG
(``RE nodes removed''; 3,332 nodes; 6,901 edges; 2,933 trees; note that msprime
uses distinct nodes for the left and right parent of each recombination).
The curve we obtain is indistinguishable from that of the full
ARG, demonstrating that the locally unary portions of coalescent nodes
contain all the information that we need about recombination events
for this application. \revpoint{2}{31}
When we fully simplify the simulated ARG
to remove all locally unary portions of nodes from trees
(``Fully simplified''; 3,332 nodes; 10,674 edges, 2,933 trees)  
we can see there is a significant loss of information, \revpoint{2}{32}
and the minimum of the
likelihood ratio surface is biased away from its true value,
as expected, because removing the unary portions of nodes has removed some of the area
of the ARG in which recombination did not occur.
(See~\cite{wong_general_2023} for more discussion on varying
degrees of ARG simplification.)
